Approximation as Differentiation: Notes

### 2025-05-27 main/category-of-families.tex

The general form of the Grothendieck construction $\Grothendieck{}T$ for an arbitrary indexed category $T:
\cat{C}^\op \to \Cat$, sometimes called the \emph{total category} for $T$, assembles all the categories $T(X)$
(for objects $X$ in $\cat{C}$) into a single category, together with morphisms that account for both internal
structure and reindexing along morphisms $f: X \to Y$ in $\cat{C}$.

@@

\noindent The $\partial X$ notation we adopt here is intended to suggest the connection between the
Grothendieck construction and the idea of tangent spaces and derivatives. To understand how the $\partial f$
components compose, consider any $(f, \partial f): (X, \partial X) \to (Y, \partial Y)$ and $(g, \partial g):
(Y, \partial Y) \to (Z, \partial Z)$ in $\Grothendieck{}T$. We have:

@@

Although reindexing is just precomposition with $f$, writing it as $\reindex{-}{f}$ avoids notational
confusion when combining composition of morphisms in $\Func{X}{C}$ with reindexing.

The Grothendieck construction for $\Func{-}{\cat{C}}: \Set^{\op} \to \Cat$, the indexed category which sends
any set $X$ to the category of $X$-indexed families $\Func{X}{\cat{C}}$ and any function $f: X \to Y$ to the
reindexing functor $\reindex{-}{f}: \Func{Y}{\cat{C}} \to \Func{X}{\cat{C}}$, is called the \emph{families}
construction for $\cat{C}$.

@@

The total category $\Fam(\LatGal)$ is then basic setting for interpreting first-order programs for \GPS. This
has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every $x \in X$, a bounded lattice
$\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of a
function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial f(x): \partial X(x) \to \partial
Y(f(x))$. This is precisely the setup we identified informally at the beginning of this section. The
reindexing along $f$ in the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial
g}{f} \comp \partial f)$ selects the appropriate lattice of approximations and Galois connection at each
point, using the baseline (unapproximated) function $f$.

>>>>

In its general form, the Grothendieck construction $\Grothendieck{}T$ for an arbitrary indexed category $T:
\cat{C}^\op \to \Cat$, sometimes called the \emph{total category} for $T$, assembles all the categories $T(X)$
(for objects $X$ in $\cat{C}$) into a single category, together with morphisms that account for both internal
structure and reindexing along morphisms $f: X \to Y$ in $\cat{C}$.

@@

\noindent The $\partial X$ notation used here is intended to suggest the connection between the Grothendieck
construction and the idea of tangent spaces and derivatives. To understand how the $\partial f$ components
compose, consider any $(f, \partial f): (X, \partial X) \to (Y, \partial Y)$ and $(g, \partial g): (Y,
\partial Y) \to (Z, \partial Z)$ in $\Grothendieck{}T$. We have:

@@

\noindent Although reindexing is just precomposition with $f$, writing it as $\reindex{-}{f}$ avoids
notational confusion when combining composition of morphisms in $\Func{X}{C}$ with reindexing.

The Grothendieck construction for $\Func{-}{\cat{C}}: \Set^{\op} \to \Cat$, the indexed category which sends
any set $X$ to the category of indexed families $\Func{X}{\cat{C}}$ and any function $f: X \to Y$ to the
reindexing functor $\reindex{-}{f}: \Func{Y}{\cat{C}} \to \Func{X}{\cat{C}}$, is called the \emph{families}
construction for $\cat{C}$.

@@

The total category $\Fam(\LatGal)$ is then a suitable setting for interpreting first-order programs for \GPS.
It has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every $x \in X$, a bounded lattice
$\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of a
function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial f(x): \partial X(x) \to \partial
Y(f(x))$. This is precisely the setup we identified informally at the beginning of this section. The
reindexing along $f$ in the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial
g}{f} \comp \partial f)$ selects the appropriate lattice of approximations and Galois connection at each
point, using the baseline (unapproximated) function $f$.

### 2025-05-27 main/formal-structures.tex

$\LatGal$ is not Cartesian closed, and so cannot serve directly as the basis for interpreting higher-order
programs. In fact Cartesian closure is incompatible with another notable property of $\LatGal$. The
projections $\pi_1: X \times Y \to X$ and $\pi_2: X \times Y \to Y$, where $X \times Y$ denotes the product of
lattices, have both upper and lower adjoints; this means that $X \times Y$, which we shall hereafter write as
$X \biprod Y$, is both a product and a coproduct, with projections $\biproj_X$ and $\biproj_Y$ and
injections $\biinj_X$ and $\biinj_Y$ given by:

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we write as $0$, is the unit for
$\biprod$. Such a structure $(\biprod, 0)$ where $\biprod$ acts as both a product and a coproduct is called a
\emph{biproduct} (\secref{biproducts}). The presence of a zero object in a Cartesian closed category $\cat{C}$
renders $\cat{C}$ trivial; if $0 \times X \iso X$ then $\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X
\Rightarrow Y} \iso 1$, and $\cat{C}$ has only a single object, up to isomorphism. However although biproducts
rule out Cartesian closure, they will prove crucial for establishing a Cartesian closed setting in
\secref{higher-order}; but let us first see how to develop a suitable categorical setting for first-order
programs.

>>>>

$\LatGal$ is not Cartesian closed, and so not directly suitable for interpreting higher-order programs. In
fact Cartesian closure is incompatible with another important property of $\LatGal$. The projections $\pi_1: X
\times Y \to X$ and $\pi_2: X \times Y \to Y$, where $X \times Y$ denotes the product of lattices, have both
upper and lower adjoints; this means that $X \times Y$, which we shall hereafter write as $X \biprod Y$, is
both a product and a coproduct, with projections $\biproj_X$ and $\biproj_Y$ and injections $\biinj_X$ and
$\biinj_Y$ given by:

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we write as $0$, is the unit for
$\biprod$. Such a structure $(\biprod, 0)$ where $\biprod$ acts as both a product and a coproduct is called a
\emph{biproduct} (\secref{biproducts}). The presence of a zero object in a Cartesian closed category $\cat{C}$
renders $\cat{C}$ trivial; if $0 \times X \iso X$ then $\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X
\Rightarrow Y} \iso 1$, and so $\cat{C}$ has only a single object, up to isomorphism. \todo{Bridge better to
next section; even if $\LatGal$ were Cartesian closed, directly interpreting into $\LatGal$ wouldn't allow
each point to be associated with its own lattice of approximants.}

### 2025-05-27 main/introduction.tex

Previous work on \GPS uses an operational semantics of programs that generates a trace of each execution which can be rewound or replayed after the fact, to compute the Galois connection described above. It would be satisfying to have a denotational account of \GPS as well, for which traces would then offer a plausible implementation technique (perhaps one of several). As it happens there is a close analogy between \GPS and \emph{automatic differentiation} for differentiable programs~\cite{siskind08,elliott18}, which points to a way to develop such an approach. We have already hinted at this in the description above, but let us now make it explicit.

@@

\begin{enumerate}
\item We explain how the CHAD framework of Vákár et al. can be adapted to give a general categorical framework for \GPS,
\item Using our adapted CHAD framework, we can explain how type structure can be used to control the approximation lattices associated with data points, something that was ``hard coded'' in previous presentations of \GPS.
\item With the benefit of our abstract setting, we can relate \GPS to other parts of denotational semantics. In particular, we show that there is a close connection with Stable Domain Theory, a proposed solution to capturing sequentiality by recording more intensional information about programs' sensitivity to approximations.
\item We also relate \GPS to categorical treatments of differentiability via Tangent Categories \note{Bob: not sure about this.}
\end{enumerate}

>>>>

Previous work on \GPS used a special operational semantics to generate a trace of each execution, and then separately used that trace to compute the Galois connection described above, by re-running forwards or backwards over the trace. It would be useful to have a denotational account of \GPS as well, especially if we could provide a semantics where the backwards analysis is baked in, rather than provided by a separately (but somewhat redundantly) defined ``backwards evaluation'' operation. As it happens there is a close analogy between \GPS and \emph{automatic differentiation} for differentiable programs~\cite{siskind08,elliott18,vákár22}, which points to a way to develop such an approach. We have already hinted at this in the description above, but let us now make it explicit.

@@

\begin{enumerate}
\item We explain how the CHAD framework of Vákár et al. can be adapted to give a general categorical framework for \GPS.
\item Using our adapted CHAD framework, we can explain how type structure can be used to control the approximation lattices associated with data points, something that was ``hard coded'' in previous presentations of \GPS.
\item With the benefit of our abstract setting, we can relate \GPS to other parts of denotational semantics. In particular, we show that there is a close connection with Stable Domain Theory, a proposed solution to capturing sequentiality by recording more intensional information about programs' sensitivity to approximations.
\item We also relate \GPS to categorical treatments of differentiability via Tangent Categories \note{Bob: not sure about this.}
\end{enumerate}
